---
title: "Spatially constrained hierarchical clustering"
---

### 1.Introduction

In today's class exercise, lets see how to apply Spatially constrained clustering algorithm - ClustGeo Method to delineate homogeneous region of Vietnam.

### 2.Data

Following two data sets are used:

-   Myanmar Township Boundary Data (i.e. *myanmar_township_boundaries*) : This is a GIS data in ESRI shapefile format. It consists of township boundary information of Myanmar. The spatial data are captured in polygon features.

-   *Shan-ICT.csv*: This is an extract of [**The 2014 Myanmar Population and Housing Census Myanmar**](https://myanmar.unfpa.org/en/publications/2014-population-and-housing-census-myanmar-data-sheet) at the township level.

[`p_load`](https://cran.r-project.org/web/packages/pacman/pacman.pdf) function pf [**pacman**](https://github.com/trinker/pacman) package is used to install and load sf all necessary packages into R environment.

-   **sf**, **rgdal** and **spdep -** Spatial data handling

-   **tidyverse**, especially **readr**, **ggplot2** and **dplyr -** Attribute data handling

-   **tmap -**Choropleth mapping

-   **coorplot**, **ggpubr**, and **heatmaply -** Multivariate data visualisation and analysis

-   **cluster, ClustGeo -** Cluster analysis

The code chunk below installs and launches these R packages into R environment.

```{r}
#| code-fold: true
#| code-summary: "Loading packages"

pacman::p_load(rgdal, spdep, tmap, sf, 
               ggpubr, cluster, factoextra, NbClust,
               heatmaply, corrplot, psych, tidyverse, purrr,ClustGeo)
```

### 3. Geospatial Analysis

#### 3.1 Importing data

The Myanmar Township Boundary GIS data is in ESRI shapefile format. It is imported into R environment by using the [*st_read()*](https://www.rdocumentation.org/packages/sf/versions/0.7-2/topics/st_read) function of **sf**. The imported InfoComm variables are extracted from **The 2014 Myanmar Population and Housing Census Myanmar**. The attribute data set is called *ict*. It is saved in R's tibble data.frame format.

```{r, message=FALSE}
#| code-fold: true
#| code-summary: "Importing data"

shan_sf <- st_read(dsn = "data3/geospatial", 
                   layer = "myanmar_township_boundaries") %>%
  filter(ST %in% c("Shan (East)", "Shan (North)", "Shan (South)"))

ict <- read_csv ("data3/aspatial/Shan-ICT.csv")

```

#### 5.2 Data Wrangling

It is wiser to take proportion of no. of households present instead of using the numbers directly. So, let us preprocess the data accordingly by using the code chunk below:

```{r}
#| code-fold: true
#| code-summary: "Preprocessing data"

ict_derived <- ict %>%
  mutate(`RADIO_PR` = `Radio`/`Total households`*1000) %>%
  mutate(`TV_PR` = `Television`/`Total households`*1000) %>%
  mutate(`LLPHONE_PR` = `Land line phone`/`Total households`*1000) %>%
  mutate(`MPHONE_PR` = `Mobile phone`/`Total households`*1000) %>%
  mutate(`COMPUTER_PR` = `Computer`/`Total households`*1000) %>%
  mutate(`INTERNET_PR` = `Internet at home`/`Total households`*1000) %>%
  rename(`DT_PCODE` =`District Pcode`,`DT`=`District Name`,
         `TS_PCODE`=`Township Pcode`, `TS`=`Township Name`,
         `TT_HOUSEHOLDS`=`Total households`,
         `RADIO`=`Radio`, `TV`=`Television`, 
         `LLPHONE`=`Land line phone`, `MPHONE`=`Mobile phone`,
         `COMPUTER`=`Computer`, `INTERNET`=`Internet at home`) 

```

#### 5.3 Visualising the distribution

```{r, fig.height=8, fig.width=12}
#| code-fold: true
#| code-summary: "Histogram for all ICT medium"

radio <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`,
                 y= ..density..)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="coral")+
  geom_density(color="black",
               alpha = 0.2)

tv <- ggplot(data=ict_derived, 
             aes(x= `TV_PR`,
                 y= ..density..)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="coral")+
  geom_density(color="red",
               alpha = 0.2)

llphone <- ggplot(data=ict_derived, 
             aes(x= `LLPHONE_PR`,
                 y= ..density..)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="coral")+
    geom_density(color="black",
               alpha = 0.2)

mphone <- ggplot(data=ict_derived, 
             aes(x= `MPHONE_PR`,
                 y= ..density..)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="coral")+
  geom_density(color="black",
               alpha = 0.2)

computer <- ggplot(data=ict_derived, 
             aes(x= `COMPUTER_PR`,
                 y= ..density..)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="coral")+
  geom_density(color="black",
               alpha = 0.2)

internet <- ggplot(data=ict_derived, 
             aes(x= `INTERNET_PR`,
                 y= ..density..)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="coral")+
  geom_density(color="black",
               alpha = 0.)

ggarrange(radio, tv, llphone, mphone, computer, internet, 
          ncol = 3, 
          nrow = 2)
```

#### 5.4 Correlation Analysis

Before we perform cluster analysis, it is important for us to ensure that the cluster variables are not highly correlated.

In this section, you will learn how to use [*corrplot.mixed()*](https://cran.r-project.org/web/packages/corrplot/corrplot.pdf) function of [**corrplot**](https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html) package to visualise and analyse the correlation of the input variables.

```{r}
#| code-fold: true
#| code-summary: "Correlation"

cluster_vars.cor = cor(ict_derived[,12:17])
corrplot.mixed(cluster_vars.cor,
         lower = "ellipse", 
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black")
```

Before we visualise the maps using choropleth, let us first combine both geospatial and aspatial datsets into one simple feature dataframe using the code chunk below.

```{r}
#| code-fold: true
#| code-summary: "Combining geospatial and aspatial data"

shan_sf <- left_join(shan_sf, 
                     ict_derived, 
                     by=c("TS_PCODE"="TS_PCODE"))
```

#### 5.5 Hierarchy Cluster Analysis

Now let us perform hierarchical cluster analysis. The analysis consists of four major steps:

1.  Extracting clustering variables

2.  Data standardisation

3.  Visualising the clustering variables

4.  Computing Proximity Matrix

**Extracting clustering variables**

```{r}
#| code-fold: true
#| code-summary: "Clustering"

cluster_vars <- shan_sf %>%
  st_set_geometry(NULL) %>%
  select("TS.x", "RADIO_PR", "TV_PR", "LLPHONE_PR", "MPHONE_PR", "COMPUTER_PR")
head(cluster_vars,10)
```

Now let us rename the township name instead of row number and delete the old one using the code chunk below

```{r}
#| code-fold: true
#| code-summary: "Renaming column"

row.names(cluster_vars) <- cluster_vars$"TS.x"
shan_ict <- select(cluster_vars, c(2:6))
head(shan_ict, 10)
```

**Data Standardisation**

In order to avoid the cluster analysis result is baised to clustering variables with large values, it is useful to standardise the input variables before performing cluster analysis.

**Performing Min-Max standardisation**

In the code chunk below, *normalize()* of [*heatmaply*](https://cran.r-project.org/web/packages/heatmaply/) package is used to stadardisation the clustering variables by using Min-Max method. The *summary()* is then used to display the summary statistics of the standardised clustering variables.

```{r}
#| code-fold: true
#| code-summary: "Min-Max standardisation"

shan_ict.std <- normalize(shan_ict)
summary(shan_ict.std)
```

**Performing Z-score standardisation**

Z-score standardisation can be performed easily by using [*scale()*](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/scale) of Base R. The code chunk below will be used to stadardisation the clustering variables by using Z-score method.

```{r}
#| code-fold: true
#| code-summary: "Z standardisation"

shan_ict.z <- scale(shan_ict)
describe(shan_ict.z)
```

We observe that mean and standard deviation of the Z-score standardised clustering variables are 0 and 1 respectively. We should take note that Z-score standardisation method should only be used if we would assume all variables come from some normal distribution.

**Visualising the standardised clustering variables**

Let us visualise the standardised clustering variables graphically in addition to summary statistics

```{r, fig.width=14}
#| code-fold: true
#| code-summary: "Visualising clustering variables"


r <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")+
  ggtitle("Before Standardisation")

shan_ict_s_df <- as.data.frame(shan_ict.std)
s <- ggplot(data=shan_ict_s_df, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Min-Max Standardisation")

shan_ict_z_df <- as.data.frame(shan_ict.z)
z <- ggplot(data=shan_ict_z_df, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Z-score Standardisation")

ggarrange(r, s, z,
          ncol = 3,
          nrow = 1)
```

We can observe that after standardisation, the variables follow normal distribution.

**Computing Proximity Matrix** The code chunk below is used to compute the proximity matrix using *euclidean* method although the function *dist()* supports six distance proximity calculations, they are: **euclidean, maximum, manhattan, canberra, binary and minkowski**.

```{r}
#| code-fold: true
#| code-summary: "Proximity matrix"

proxmat <- dist(shan_ict, method = 'euclidean')

```

#### 5.6 Spatially Constrained Clustering: ClustGeo Method

#### 5.6.1 Ward-like hierarchical clustering: ClustGeo

To perform non-spatially constrained hierarchical clustering, we only need to provide the function a dissimilarity matrix as shown in the code chunk below.

```{r}
nongeo_cluster <- hclustgeo(proxmat)
plot(nongeo_cluster, cex = 0.5)
rect.hclust(nongeo_cluster, 
            k = 6, 
            border = 2:5)
```
#### 5.6.2 Mapping the clusters formed

Similarly, we can plot the clusters on a categorical area shaded map by using the steps we learned in 5.7.12 Mapping the clusters formed.

```{r}
groups <- as.factor(cutree(nongeo_cluster, k=6))
shan_sf_ngeo_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER` = `as.matrix.groups.`)
qtm(shan_sf_ngeo_cluster, "CLUSTER")
```
#### 5.6.3 Spatially Constrained Hierarchical Clustering

Before we can performed spatially constrained hierarchical clustering, a spatial distance matrix will be derived by using st_distance() of sf package.

```{r}
dist <- st_distance(shan_sf, shan_sf)
distmat <- as.dist(dist)
cr <- choicealpha(proxmat, distmat, range.alpha = seq(0, 1, 0.1), K=6, graph = TRUE)
```
With reference to the graphs above, alpha = 0.3 will be used as shown in the code chunk below

```{r}
clustG <- hclustgeo(proxmat, distmat, alpha = 0.3)
groups <- as.factor(cutree(clustG, k=6))
shan_sf_Gcluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER` = `as.matrix.groups.`)
qtm(shan_sf_Gcluster, "CLUSTER")
```



