---
title: "Spatial Patterns of Non-functional points - Nigeria"
---

#### Loading packages

[`p_load`](https://cran.r-project.org/web/packages/pacman/pacman.pdf) function pf [**pacman**](https://github.com/trinker/pacman) package is used to install and load sf and tidyverse pacages into R environment.

```{r}
pacman::p_load(sf, tidyverse,spdep, tmap, funModeling)
```

#### Importing gesospatial data

The code chunk below uses *st_read()* function of **sf** package to import *geoBoundaries-NGA-ADM2_simplified* shapefile into R as a polygon feature data frame.

Two arguments are used :

-   dsn - destination : to define the data path

-   *layer* - to provide the shapefile name

```{r}
ngaGeo <- st_read(dsn = "data2/geospatial", 
                layer = "geoBoundaries-NGA-ADM2")
```

The message above reveals that the geospatial objects are multipolygon features. There are a total of 774 multipolygon features and 5 fields in ngaGeo simple feature data frame. ngaGeo is in **wgs84** coordinates systems

#### Importing aspatial data

Since *nigeria_waterpoints* data set is in csv file format, we will use [*read_csv()*](https://readr.tidyverse.org/reference/read_delim.html) of **readr** package to import *nigeria_waterpoints.csv* as shown the code chunk below. The output R object is called *ngawaterpoints* and it is a [tibble data frame](https://r4ds.had.co.nz/tibbles.html).

```{r}
ngawaterpoints  <- read_csv("data2/aspatial/nigeria_waterpoints.csv")
```

```{r}
list(ngawaterpoints)
```

#### Creating a simple feature data frame from an aspatial data frame

The code chunk below converts *ngawaterpoints* data frame into a simple feature data frame by using [*st_as_sf()*](https://r-spatial.github.io/sf/reference/st_as_sf.html) of **sf** packages

```{r}
ngawater_sf <- st_as_sf(ngawaterpoints , 
                       coords = c("#lon_deg", "#lat_deg"),
                       crs=4326) %>%
  st_transform(crs = 26391)
```

#### Data Wrangling

```{r}
ngawater_sf <-ngawater_sf %>% 
mutate(waterpoint_status=recode(`#status_clean`, 
                     'Abandoned'='Non-Functional',
                     'Abandoned/Decommissioned'='Non-Functional',
                     'Non functional due to dry season'='Non-Functional',
                     'Non-Functional due to dry season'='Non-Functional',
                     'Functional but not in use'='Functional',
                     'Functional but needs repair'= 'Functional'))
```

```{r}
unique(ngawater_sf$waterpoint_status)
```
