---
title: "Geospatial Data Wrangling"
author: "Raveena Chakrapani"
---

### data

```{r}
pacman::p_load(sf, tidyverse, tmap)
```

```{r}
mpsz = st_read(dsn = "data/geospatial", 
                  layer = "MP14_SUBZONE_WEB_PL")
```
importing polyline feature

```{r}
cyclingpath = st_read(dsn ="data/geospatial", 
                         layer = "CyclingPath")
```
### importing GIS

```{r}
preschool = st_read("data/geospatial/pre-schools-location-kml.kml")
```

```{r}
st_geometry(mpsz)
```

plotting map

```{r}
plot(mpsz)
```

plot geometry

```{r}
plot(st_geometry(mpsz))
```

plot palnning area

```{r}
plot(mpsz["PLN_AREA_N"])
```

## Working with Projection

```{r}
st_crs(mpsz)
mpsz3414 <- st_set_crs(mpsz, 3414)
st_crs(mpsz3414)
```

### Transforming the projection of preschool from wgs84 to svy21

```{r}
preschool3414 <- st_transform(preschool, 
                              crs = 3414)
```

importing aspatial data

```{r}
listings <- read_csv("data/aspatial/listings.csv")
```

dataframe from aspatial df

```{r}
listings_sf <- st_as_sf(listings, 
                       coords = c("longitude", "latitude"),
                       crs=4326) %>%
  st_transform(crs = 3414)
```

### Geoprocessing with sf package

Buffering

```{r}
buffer_cycling <- st_buffer(cyclingpath, 
                               dist=5, nQuadSegs = 30)
buffer_cycling$AREA <- st_area(buffer_cycling)
sum(buffer_cycling$AREA)
```

### point in polygon count

```{r}
mpsz3414$`PreSch Count`<- lengths(st_intersects(mpsz3414, preschool3414))
summary(mpsz3414$`PreSch Count`)
top_n(mpsz3414, 1, `PreSch Count`)
```

### density of preschool by planning subzone

```{r}
mpsz3414$Area <- mpsz3414 %>%
  st_area()
mpsz3414 <- mpsz3414 %>%
  mutate(`PreSch Density` = `PreSch Count`/Area * 1000000)
```

### Exploratory Data Analysis

```{r}
hist(mpsz3414$`PreSch Density`)
ggplot(data=mpsz3414, 
       aes(x= as.numeric(`PreSch Density`)))+
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  labs(title = "Are pre-school even distributed in Singapore?",
       subtitle= "There are many planning sub-zones with a single pre-school, on the other hand, \nthere are two planning sub-zones with at least 20 pre-schools",
      x = "Pre-school density (per km sq)",
      y = "Frequency")
```